---
title: "TSOpt walkthrough: every function, from data upload to training set"
subtitle: "A step-by-step check of the whole package on your own data, in every breeding system"
author: "PlantNura Consulting & Solutions"
date: "`r format(Sys.Date(), '%d %B %Y')` · TSOpt `r as.character(utils::packageVersion('TSOpt'))`"
params:
  geno: ""
  pheno: ""
  truth: ""
  groups: ""
  costs: ""
  effects: ""
  history: ""
  history_trait: "y"
  cm_per_mb: -1
  omics: ""
  omics_type: "auto"
  n_train: 60
  h2: 0.5
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
    dev: ragg_png
    fig_width: 7
    fig_height: 4.2
    fig_crop: false
    includes:
      in_header: walkthrough_style.tex
mainfont: DejaVuSans.ttf
mainfontoptions:
  - BoldFont=DejaVuSans-Bold.ttf
  - ItalicFont=DejaVuSans-Oblique.ttf
  - BoldItalicFont=DejaVuSans-BoldOblique.ttf
monofont: DejaVuSansMono.ttf
monofontoptions:
  - BoldFont=DejaVuSansMono-Bold.ttf
  - ItalicFont=DejaVuSansMono-Oblique.ttf
  - BoldItalicFont=DejaVuSansMono-BoldOblique.ttf
fontsize: 10pt
geometry: margin=2.2cm
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
                      dpi = 200, out.width = "100%", fig.align = "center")
options(width = 92, digits = 4)
```

# How to use this document {-}

This document runs **every exported function of TSOpt** in the order a breeder uses them:
upload the genotype file, check its quality, build the relationship matrices, pack everything
into one data object, ask whether optimisation is worth it, try every criterion and every
search, build training sets for every breeding system, protect major genes, check the plans,
and plan the next season. Each step prints its result and a short explanation of what to look
for.

**Run it on your own data.** The document has parameters. In RStudio, open it and choose
*Knit* > *Knit with Parameters...*, then fill in the paths to your files. Or render it from the
console:

```{r how-to-run, eval = FALSE}
rmarkdown::render("docs/walkthrough/TSOpt_Walkthrough.Rmd",
                  params = list(geno   = "my_panel.vcf.gz",    # any format tso_read() reads
                                pheno  = "my_pheno.csv",       # columns: id, y
                                groups = "my_groups.csv",      # columns: id, group (optional)
                                costs  = "my_costs.csv",       # columns: id, cost  (optional)
                                effects = "my_effects.csv",    # marker, effect[, allele, p] (optional)
                                history = "my_history.csv",    # id, cycle, trait: forward validation (optional)
                                cm_per_mb = 0.5,               # map density, for cross prediction (optional)
                                omics = "my_profiles.csv",     # id, one column per feature/wavelength (optional)
                                n_train = 100, h2 = 0.4))
```

Every parameter left empty falls back to the example data shipped with TSOpt: a simulated
panel of 300 inbred lines and 1,000 SNPs, with phenotypes, true genetic values, groups and
per-line costs. Because the example is simulated, the true genetic values are known, so
Section 12 can measure how good a training set really was.

**Your licence key must be active** (`tso_licence("TSO1....")` once on this computer;
`tso_licence()` shows it). TSOpt is free for everyone; request a key with
`tso_licence_request()`.

**Run it one chunk at a time.** Put the cursor in a chunk and press Ctrl/Cmd + Shift + Enter.
The chunks depend on each other in order, so run them from the top.

**Spot anything odd.** After almost every function there is an automatic check, printed as
`[PASS]` or `[FAIL]`. It tests something that must hold if the function worked: the relationship
matrix is symmetric, no target line ends up in its own training set, a budget is respected,
and so on. Section 16 collects every check into a scorecard. A `[FAIL]` is either a bug in TSOpt
or a problem with the input data, and the check's text says which quantity to look at.

\newpage

# Set-up

```{r libraries}
library(TSOpt)
library(ggplot2)
packageVersion("TSOpt")

ex <- function(f) system.file("extdata", f, package = "TSOpt")   # shipped example files
pick <- function(p, default) if (nzchar(p)) p else default
geno_file  <- pick(params$geno,   ex("example_panel.vcf.gz"))
pheno_file <- pick(params$pheno,  ex("example_pheno.csv"))
truth_file <- pick(params$truth,  if (nzchar(params$geno)) "" else ex("example_true_values.csv"))
group_file <- pick(params$groups, if (nzchar(params$geno)) "" else ex("example_groups.csv"))
cost_file  <- pick(params$costs,  if (nzchar(params$geno)) "" else ex("example_costs.csv"))
eff_file   <- pick(params$effects, if (nzchar(params$geno)) "" else ex("example_marker_effects.csv"))
using_example <- !nzchar(params$geno)
out <- file.path(tempdir(), "tsopt_walkthrough"); dir.create(out, showWarnings = FALSE)
cat("Genotypes:", geno_file, "\nPhenotypes:", pheno_file,
    "\nUsing the shipped example data:", using_example, "\n")
```

The scorecard helper records every check:

```{r checks}
checks <- data.frame(section = character(), check = character(), result = character())
section <- "Set-up"
chk <- function(what, ok, detail = "") {
  ok <- isTRUE(ok)
  checks[nrow(checks) + 1, ] <<- list(section, what, if (ok) "PASS" else "FAIL")
  cat(if (ok) "[PASS] " else "[FAIL] ", what, if (nzchar(detail)) paste0("  (", detail, ")"), "\n",
      sep = "")
  invisible(ok)
}
timed <- function(expr) { t0 <- proc.time()[["elapsed"]]; v <- expr
  attr(v, "seconds") <- proc.time()[["elapsed"]] - t0; v }
```

\newpage

# Uploading genotype data

```{r s1, include = FALSE}
section <- "1 Upload"
```

## Your file

`tso_read()` detects the file format, the genotype coding (0/1/2, -1/0/1, letters, IUPAC,
AB calls, dosages), the missing-value codes, the ploidy and whether markers are in rows or
columns. It records every decision in a log.

```{r read-main}
g_raw <- tso_read(geno_file)
g_raw
```

What to look for: the number of lines and markers must match what your genotyping provider
delivered, the detected coding must be right, and the missing rate should be plausible (a
few per cent for arrays, often more for GBS).

```{r read-main-checks}
D <- g_raw$dosage
chk("dosage matrix has lines in rows and markers in columns", nrow(D) > 1 && ncol(D) > 1,
    paste(nrow(D), "lines x", ncol(D), "markers"))
chk("line IDs are unique", !anyDuplicated(rownames(D)))
chk("marker IDs are unique", !anyDuplicated(colnames(D)))
chk("dosages lie between 0 and the ploidy", all(D >= 0 & D <= g_raw$ploidy, na.rm = TRUE),
    paste("ploidy", g_raw$ploidy))
chk("missing rate below 50%", mean(is.na(D)) < 0.5, paste0(round(100 * mean(is.na(D)), 2), "% missing"))
```

## Every format TSOpt reads

The shipped example panel exists in five formats. Reading all of them must give the same
lines and markers, and the same relationship matrix (allele coding may flip between formats,
but relationships must not change).

```{r read-formats}
fmt <- list(VCF    = ex("example_panel.vcf.gz"),
            HapMap = ex("example_panel.hmp.txt.gz"),
            CSV    = ex("example_panel.csv.gz"),
            PLINK  = sub("\\.bed$", "", ex("example_plink.bed")))
gs <- lapply(fmt, tso_read, verbose = FALSE)
data.frame(format = names(gs), lines = sapply(gs, function(x) nrow(x$dosage)),
           markers = sapply(gs, function(x) ncol(x$dosage)),
           coding = sapply(gs, function(x) x$format), row.names = NULL)
K_by_format <- lapply(gs, tso_kinship, scale = FALSE)
diffs <- sapply(K_by_format[-1], function(K) max(abs(K - K_by_format$VCF)))
round(diffs, 12)
chk("VCF, HapMap, CSV and PLINK give the same relationship matrix", all(diffs < 1e-8))
g_dart <- tso_read(ex("example_dart.csv"), verbose = FALSE)
chk("DArTseq one-row file is read", nrow(g_dart$dosage) > 0,
    paste(nrow(g_dart$dosage), "lines x", ncol(g_dart$dosage), "markers"))
```

## Writing and converting

`tso_write()` converts between formats; the extension chooses the format. Every written file
is read back and must give the same relationship matrix as the original.

```{r write-formats}
exts <- c(".vcf", ".vcf.gz", ".hmp.txt", ".bed", ".csv", ".raw", ".rds")
K0 <- tso_kinship(gs$VCF, scale = FALSE)
rt <- sapply(exts, function(e) {
  f <- file.path(out, paste0("roundtrip", e))
  tso_write(gs$VCF, f)
  back <- tso_read(if (e == ".bed") sub("\\.bed$", "", f) else f, verbose = FALSE)
  max(abs(tso_kinship(back, scale = FALSE) - K0))
})
round(rt, 12)
chk("every writable format round-trips to the same kinship", all(rt < 1e-8))
if (requireNamespace("writexl", quietly = TRUE) && requireNamespace("readxl", quietly = TRUE)) {
  xl <- file.path(out, "panel.xlsx")
  writexl::write_xlsx(data.frame(id = rownames(gs$VCF$dosage), gs$VCF$dosage, check.names = FALSE), xl)
  g_xl <- tso_read(xl, verbose = FALSE)
  chk("an Excel sheet is read", identical(dim(g_xl$dosage), dim(gs$VCF$dosage)))
}
```

## Unusual inputs

Real files are messy. These small inputs test the codings breeders meet most often: allele
letters in several notations, a matrix with markers in rows, Axiom-style -1 for missing, and
an autotetraploid dosage matrix.

```{r odd-inputs}
ids5 <- paste0("L", 1:5)
calls <- list(pairs = c("AA", "AG", "GG", "AA", "NN"), slash = c("A/A", "A/G", "G/G", "A/A", "./."),
              iupac = c("A", "R", "G", "A", "N"), ab = c("AA", "AB", "BB", "AA", "--"))
dos <- sapply(calls, function(v) as.vector(tso_read(matrix(v, 5, 1, dimnames = list(ids5, "m1")),
                                                    markers_in = "columns", verbose = FALSE)$dosage))
dos
chk("allele pairs, slashes, IUPAC and AB calls give the same dosages",
    all(apply(dos, 1, function(r) length(unique(r)) == 1), na.rm = TRUE))
Xs <- gs$VCF$dosage[1:20, 1:40]
rows_df <- data.frame(marker = colnames(Xs), t(Xs), check.names = FALSE)
chk("markers in rows are transposed automatically",
    identical(rownames(tso_read(rows_df, verbose = FALSE)$dosage), rownames(Xs)))
ax <- Xs - 1; ax[1, 1] <- -9
chk("-1/0/1 coding is recognised", isTRUE(all.equal(unname(tso_read(Xs - 1, verbose = FALSE)$dosage), unname(Xs))))
set.seed(4)
tet <- matrix(sample(0:4, 30 * 50, TRUE), 30, 50, dimnames = list(paste0("C", 1:30), paste0("m", 1:50)))
chk("0-4 dosages are read as an autotetraploid", tso_read(tet, verbose = FALSE)$ploidy == 4)
```

\newpage

# Quality control

```{r s2, include = FALSE}
section <- "2 Quality control"
```

`tso_qc()` removes badly called lines, then markers with a low call rate or a low minor-allele
frequency, imputes what is left missing, and stores the base allele frequencies. Every step goes
into the log.

```{r qc-default}
g <- tso_qc(g_raw)
```

```{r qc-checks}
chk("QC keeps at least half of the markers", ncol(g$dosage) >= 0.5 * ncol(g_raw$dosage),
    paste(ncol(g$dosage), "of", ncol(g_raw$dosage)))
chk("no missing values left after imputation", !anyNA(g$dosage))
mafs <- pmin(g$base_freq, 1 - g$base_freq)
chk("every kept marker has MAF >= 0.05", all(mafs >= 0.05 - 1e-12))
chk("base allele frequencies are stored, one per marker", length(g$base_freq) == ncol(g$dosage))
chk("counted alleles are stored with the base frequencies",
    !is.null(attr(g$base_freq, "counted")) || is.null(g_raw$map$counted))
```

The distributions behind these decisions:

```{r qc-plots, fig.height = 3.2}
qc_df <- rbind(data.frame(what = "minor-allele frequency (kept markers)", value = mafs),
               data.frame(what = "call rate per marker (before QC)", value = colMeans(!is.na(g_raw$dosage))))
ggplot(qc_df, aes(value)) + geom_histogram(bins = 30, fill = tso_colors()[["training"]]) +
  facet_wrap(~ what, scales = "free") + tso_theme() + labs(x = NULL, y = "markers")
```

A stricter set of thresholds, for comparison:

```{r qc-strict}
g_strict <- tso_qc(g_raw, maf = 0.10, call_rate = 0.95, line_call_rate = 0.9, verbose = FALSE)
c(default = ncol(g$dosage), strict = ncol(g_strict$dosage))
chk("stricter thresholds keep fewer (or equal) markers", ncol(g_strict$dosage) <= ncol(g$dosage))
```

## Allele orientation across breeding cycles

HapMap and allele-letter files are coded by each file's own minor allele, so the same marker can
flip between two files. When the base frequencies of an earlier cycle are reused, TSOpt re-orients
the new file to the stored counted allele.

```{r orientation}
if (using_example) {
  hmp_sub <- gs$HapMap; hmp_sub$dosage <- hmp_sub$dosage[1:60, ]
  bf <- tso_qc(gs$VCF, verbose = FALSE)$base_freq
  aligned <- tso_qc(hmp_sub, base_freq = bf, verbose = FALSE)
  flipped <- grep("re-oriented", aligned$log, value = TRUE)
  cat(if (length(flipped)) flipped else "no marker needed re-orienting", "\n")
  vcf_qc <- tso_qc(gs$VCF, verbose = FALSE)
  cm <- intersect(colnames(aligned$dosage), colnames(vcf_qc$dosage))
  called <- !is.na(gs$VCF$dosage[1:60, cm]) & !is.na(gs$HapMap$dosage[1:60, cm])
  agree <- mean((aligned$dosage[, cm] == vcf_qc$dosage[1:60, cm])[called])
  cat("re-oriented HapMap dosages agree with the VCF on", round(100 * agree, 2), "% of called genotypes\n")
  chk("re-oriented HapMap dosages equal the VCF dosages", agree == 1)
  # K is rescaled to mean diagonal 1 within each panel, so compare the unscaled matrices
  sub_vcf <- vcf_qc; sub_vcf$dosage <- vcf_qc$dosage[1:60, cm]
  chk("and give the same (unscaled) relationship matrix",
      max(abs(tso_kinship(aligned, base_freq = bf[cm], scale = FALSE) -
              tso_kinship(sub_vcf, base_freq = bf[cm], scale = FALSE))) < 1e-8)
}
```

\newpage

# Relationship matrices

```{r s3, include = FALSE}
section <- "3 Relationship matrices"
```

## Additive (VanRaden)

```{r kin-add}
K <- tso_kinship(g)
round(K[1:6, 1:6], 3)
ev <- eigen(K, symmetric = TRUE, only.values = TRUE)$values
c(mean_diagonal = mean(diag(K)), min_eigenvalue = min(ev), mean_off_diagonal = mean(K[upper.tri(K)]))
chk("K is symmetric", isSymmetric(unname(K)))
chk("K is positive semi-definite", min(ev) > -1e-8)
chk("mean diagonal of K is 1 (entry-mean scale)", abs(mean(diag(K)) - 1) < 1e-3)
chk("row and column names are the line IDs", identical(rownames(K), rownames(g$dosage)))
```

The eigenvalue spectrum tells how structured the panel is: a few dominant eigenvalues mean
strong population structure; a flat spectrum means a continuum.

```{r kin-spectrum, fig.height = 3}
ggplot(data.frame(k = seq_along(ev)[1:30], share = (ev / sum(ev))[1:30]), aes(k, share)) +
  geom_col(fill = tso_colors()[["training"]]) + tso_theme() +
  labs(x = "principal component", y = "share of genetic variance", title = "Spectrum of K")
```

The relationship matrix as a heat map, lines ordered by genetic group (each block on the
diagonal is a subpopulation or family):

```{r kin-heatmap, fig.height = 5}
tso_plot(tso_data(g, groups = if (nzchar(group_file)) {
  d <- utils::read.csv(group_file); stats::setNames(d$group, as.character(d$id)) }, qc = FALSE,
  verbose = FALSE), "kinship")
```

## Dominance, weighted and pedigree kernels

```{r kin-other}
if (g$ploidy == 2) {
  Kd <- tso_kinship(g, method = "dominance")
  chk("dominance kernel is symmetric and PSD",
      isSymmetric(unname(Kd)) && min(eigen(Kd, TRUE, only.values = TRUE)$values) > -1e-8)
}
w <- setNames(rep(1, ncol(g$dosage)), colnames(g$dosage)); w[1:50] <- 10
Kw <- tso_kinship(g, weights = w)
chk("weighted kinship differs from the plain one", max(abs(Kw - K)) > 1e-4)
w0 <- w; w0[] <- 1
chk("weights all equal to 1 reproduce the plain kinship", max(abs(tso_kinship(g, weights = w0) - K)) < 1e-10)
ped <- data.frame(id = c("F1", "F2", "F3", "S1", "S2", "S3", "G1"),
                  sire = c(NA, NA, NA, "F1", "F1", "F1", "S1"),
                  dam  = c(NA, NA, NA, "F2", "F2", "F3", "S3"))
A <- tso_kinship(method = "pedigree", pedigree = ped, scale = FALSE, ridge = 0)
A
chk("pedigree: full sibs S1, S2 are related 0.5", abs(A["S1", "S2"] - 0.5) < 1e-12)
chk("pedigree: half sibs S1, S3 are related 0.25", abs(A["S1", "S3"] - 0.25) < 1e-12)
chk("pedigree: G1 (S1 x S3) is inbred, A = 1.125", abs(A["G1", "G1"] - 1.125) < 1e-12)
```

## Epistatic, non-parametric and single-step kernels

`tso_kernels()` builds every covariance model the design can use. Dominance kernels are only
built when your material is heterozygous (inbred lines carry no dominance):

```{r kernels}
het <- mean(abs(g$dosage - 1) < 0.5, na.rm = TRUE)
which_k <- c("A", if (g$ploidy == 2 && het > 0.01) "D", "AA", "gaussian")
kk <- tso_kernels(g, which = which_k)
kk
chk("every kernel has mean diagonal 1", all(abs(sapply(kk$K, function(M) mean(diag(M))) - 1) < 1e-10))
chk("every kernel is positive semi-definite",
    all(sapply(kk$K, function(M) min(eigen(M, TRUE, only.values = TRUE)$values)) > -1e-8))
AA0 <- kk$K$A * kk$K$A
chk("AA is the element-wise square of A", max(abs(kk$K$AA - AA0 / mean(diag(AA0)))) < 1e-8)
off <- upper.tri(kk$K$A)
chk("epistatic relationships fall off faster than additive ones",
    mean(abs(kk$K$AA[off])) < mean(abs(kk$K$A[off])))
Dp <- TSOpt:::.pedigree_D(ped, TSOpt:::.pedigree_A(ped))
chk("pedigree dominance: full sibs share 0.25, half sibs 0", abs(Dp["S1", "S2"] - 0.25) < 1e-12 && Dp["S1", "S3"] == 0)
# single-step H: 10 new crosses of genotyped parents, not genotyped themselves
par_ids <- rownames(g$dosage)[1:20]
ped_h <- data.frame(id = c(par_ids, paste0("new", 1:10)), p1 = c(rep(NA, 20), par_ids[1:10]),
                    p2 = c(rep(NA, 20), par_ids[11:20]))
gH <- g; gH$dosage <- g$dosage[par_ids, ]
H <- tso_kernels(gH, pedigree = ped_h, which = "H")$K$H
chk("H covers the ungenotyped crosses", all(paste0("new", 1:10) %in% rownames(H)))
chk("a new cross is closer to its parent than to other lines", H["new1", par_ids[1]] > mean(H["new1", par_ids[-c(1, 11)]]))
tp_k <- tso_data(g, kernels = kk, kernel_share = "robust", qc = FALSE, verbose = FALSE)
chk("tso_data combines the kernels at the centre shares",
    max(abs(tp_k$K - TSOpt:::.combine_kernels(tp_k$kernels$K, tp_k$kernels$share))) < 1e-10)
```

\newpage

# The data object and the breeding systems

```{r s4, include = FALSE}
section <- "4 Data object"
```

`tso_data()` packs genotypes, the relationship matrix, the breeding system, groups, families and
costs into one object. A *unit* is anything you can phenotype: a line, a hybrid, a clone, a line
in one environment, or one trait on one line.

## Inbred lines (wheat, rice, barley, soybean)

```{r data-inbred}
read_side <- function(f, col) if (nzchar(f) && file.exists(f)) { d <- utils::read.csv(f)
  stats::setNames(d[[col]], as.character(d$id)) } else NULL
groups <- read_side(group_file, "group")
costs  <- read_side(cost_file, "cost")
tp <- tso_data(g, groups = groups, cost = costs, qc = FALSE)
ids <- tp$units$id; N <- length(ids)
n_train <- min(params$n_train, floor(0.3 * N))
targets <- ids[seq(N - round(0.15 * N) + 1, N)]    # the last 15% of lines play "this season"
cat("units:", N, " training-set size used below:", n_train, " targets:", length(targets), "\n")
chk("one unit per line", nrow(tp$units) == nrow(g$dosage))
chk("the kernel covers every unit", identical(dim(tp$K), c(N, N)))
if (!is.null(costs)) chk("costs attached to every unit", !anyNA(tp$units$cost))
```

The population structure the design will see:

```{r data-pca, fig.height = 4}
tso_plot(tp)
```

## Clonal crops: diploid (cassava, banana) and tetraploid (potato)

For clones the non-additive part matters, so a dominance kernel is added for diploids. The
tetraploid panel is built from two outbred diploid simulations (dosages 0 to 4).

```{r data-clonal}
tp_clone <- tso_data(g, system = "clonal", qc = FALSE, verbose = FALSE)
chk("clonal kernel differs from the additive one", max(abs(tp_clone$K - tp$K)) > 1e-6)
p_a <- tso_simulate_panel(n = 150, markers = 600, seed = 31, type = "outbred")
p_b <- tso_simulate_panel(n = 150, markers = 600, seed = 32, type = "outbred")
X4 <- p_a$dosage + p_b$dosage; rownames(X4) <- paste0("POT", seq_len(nrow(X4)))
g_potato <- tso_read(X4, ploidy = 4, verbose = FALSE)
tp_potato <- tso_data(g_potato, system = "clonal")
chk("tetraploid ploidy detected", g_potato$ploidy == 4)
```

## Outcrossing crops (ryegrass, alfalfa, oil palm): families

```{r data-outcross}
fam <- setNames(paste0("F", ceiling(seq_len(N) / 10)), ids)
tp_out <- tso_data(g, system = "outcross", families = fam, qc = FALSE, verbose = FALSE)
chk("families attached", length(unique(tp_out$units$family)) == length(unique(fam)))
```

## Hybrids (maize, sorghum, rice hybrids)

Units are all crosses between a female and a male pool; their relationships are built from the
parents (GCA of each pool plus SCA). Hybrids involving parents not declared `crossable` can be
targets but not training units.

```{r data-hybrid}
n_pool <- min(30, floor(N / 4))
fem <- ids[1:n_pool]; mal <- ids[(n_pool + 1):(2 * n_pool)]
tp_hyb <- tso_data(g, system = "hybrid", pools = list(fem, mal), qc = FALSE,
                   crossable = c(fem[1:(n_pool - 5)], mal[1:(n_pool - 5)]))
chk("every female x male cross is a unit", nrow(tp_hyb$units) == n_pool^2)
chk("hybrids of non-crossable parents are not candidates", sum(!tp_hyb$units$crossable) > 0)
```

## Genebank or diversity panel

```{r data-genebank}
tp_bank <- tso_data(g, system = "genebank", qc = FALSE, verbose = FALSE)
chk("genebank object built", inherits(tp_bank, "tso_data"))
```

## Pedigree or kinship only

When there are no markers, a pedigree or a relationship matrix from elsewhere is enough. A
matrix that is not positive semi-definite (bent pedigrees, edited imports) is repaired with a
warning.

```{r data-kinship-only}
tp_ped <- tso_data(pedigree = ped)
tp_ped
Kbad <- K[1:40, 1:40]; Kbad[1, 2] <- Kbad[2, 1] <- 5
bent <- tryCatch(tso_data(kinship = Kbad, verbose = FALSE), warning = function(w) conditionMessage(w))
cat(substr(bent, 1, 110), "...\n")
chk("a non-PSD kinship is detected and bent", grepl("positive semi-definite", bent))
```

## Multi-environment and multi-trait data

```{r data-met-mt}
Om <- matrix(c(1, .7, .4, .7, 1, .5, .4, .5, 1), 3, dimnames = rep(list(c("Site1", "Site2", "Site3")), 2))
tp_met <- tso_data(g, envs = colnames(Om), env_cor = Om, qc = FALSE,
                   env_noise = c(Site1 = 1, Site2 = 2, Site3 = 1))
chk("units are line x environment cells", nrow(tp_met$units) == 3 * N)
tp_mt <- tso_data(g, traits = c("yield", "protein", "ndvi"), trait_h2 = c(0.3, 0.5, 0.6),
                  gcor = matrix(c(1, .4, .2, .4, 1, .1, .2, .1, 1), 3),
                  trait_cost = c(yield = 20, protein = 8, ndvi = 1), plot_cost = 10, qc = FALSE,
                  verbose = FALSE)
chk("units are line x trait cells", nrow(tp_mt$units) == 3 * N)
```

\newpage

# Is optimisation worth it here?

```{r s5, include = FALSE}
section <- "5 Diagnose and size"
```

## tso_diagnose()

```{r diagnose}
dg <- tso_diagnose(tp, n = n_train)
dg
chk("diagnosis gives a traffic-light verdict", dg$verdict$light %in% c("GREEN", "AMBER", "RED"))
chk("optimised accuracy >= random at every size", all(dg$curve$optimised_acc >= dg$curve$random_acc - 1e-9))
```

```{r diagnose-plot, fig.height = 5}
tso_plot(dg)
```

The two panels separately. The size curve shows accuracy against the number of lines
phenotyped, for optimised and random sets; the gap between the curves is what optimisation
buys. The structure panel shows the groups the stratification will use.

```{r diagnose-parts, fig.height = 3.6}
tso_plot(dg, "size_curve")
tso_plot(dg, "structure")
```

```{r diagnose-table}
knitr::kable(dg$curve, digits = 3, row.names = FALSE)
```

## tso_sample_size(): how many lines?

```{r sample-size}
ss <- tso_sample_size(tp, target = targets)
ss
chk("n for 95% <= n for 99%", ss$n_for[["0.95"]] <= ss$n_for[["0.99"]])
chk("the optimised curve never needs more plots than random", all(ss$savings >= 0, na.rm = TRUE))
```

```{r sample-size-plot}
tso_plot(ss)
```

\newpage

# Every criterion

```{r s6, include = FALSE}
section <- "6 Criteria"
```

TSOpt has 18 built-in criteria in five tiers, plus any function you write:

```{r criteria-table}
crit <- tso_criteria()
knitr::kable(crit[, c("criterion", "tier", "search")], row.names = FALSE)
```

Each criterion now designs a training set for the same problem (predict the targets, choose
`n_train` lines). Every set is then scored with the same yardstick, the exact CDmean at the
given heritability, so the table shows what each criterion trades away.

```{r criteria-loop}
crit_sets <- list()
run_crit <- function(cr) {
  d <- if (cr == "cdpop") tp_out else tp           # CDpop compares family means: families below
  p <- timed(tso_design(d, n = n_train, target = targets, criterion = cr, h2 = params$h2,
                        iterations = 100, n_random = 20, verbose = FALSE))
  crit_sets[[cr]] <<- p$selected$id
  data.frame(criterion = cr, lines = nrow(p$selected),
             targets_in_training = sum(p$selected$id %in% targets),
             exact_CDmean = tso_score(tp, train = p$selected$id, target = targets, h2 = params$h2,
                                      criteria = "cdmean_exact")[["cdmean_exact"]],
             seconds = round(attr(p, "seconds"), 2))
}
crit_res <- do.call(rbind, lapply(crit$criterion, run_crit))
crit_res$exact_CDmean <- round(crit_res$exact_CDmean, 4)
knitr::kable(crit_res[order(-crit_res$exact_CDmean), ], row.names = FALSE)
chk("every criterion returns n_train lines", all(crit_res$lines == n_train))
chk("no criterion puts a target in the training set", all(crit_res$targets_in_training == 0))
chk("cdmean (the default) beats random sampling",
    crit_res$exact_CDmean[crit_res$criterion == "cdmean"] > crit_res$exact_CDmean[crit_res$criterion == "random"])
```

```{r criteria-plot, fig.height = 4.4}
crit_res$tier <- crit$tier[match(crit_res$criterion, crit$criterion)]
ggplot(crit_res, aes(stats::reorder(criterion, exact_CDmean), exact_CDmean, fill = tier)) +
  geom_col() + coord_flip() + tso_theme() +
  scale_fill_manual(values = unname(tso_colors("groups", length(unique(crit_res$tier))))) +
  labs(x = NULL, y = "exact CDmean of the targets", fill = NULL,
       title = "Every criterion on the same problem", subtitle = "higher is better")
```

Where the training sets of four contrasting criteria sit in the genetic space:

```{r criteria-pca, fig.height = 5.5}
pcs <- tp$pca$x[, 1:2]; colnames(pcs) <- c("PC1", "PC2")
show_c <- intersect(c("cdmean", "coverage", "maximin", "kinship"), names(crit_sets))
pc_df <- do.call(rbind, lapply(show_c, function(cr) data.frame(pcs, criterion = cr,
  role = ifelse(ids %in% crit_sets[[cr]], "training", ifelse(ids %in% targets, "target", "candidate")))))
ggplot(pc_df, aes(PC1, PC2, colour = role)) + geom_point(size = 0.9, alpha = 0.8) +
  facet_wrap(~ criterion) + tso_theme() +
  scale_colour_manual(values = c(training = tso_colors()[["training"]], target = tso_colors()[["target"]],
                                 candidate = tso_colors()[["candidate"]])) + labs(colour = NULL)
```

What to look at: the CDmean-family criteria sit at the top because the yardstick is CDmean;
the model-free ones (coverage, maximin, PAM) and the baselines show how much a model-based
design adds on this panel. The `kinship` baseline (the lines most related to the targets) is
kept only for comparison: in the head-to-head benchmark it was worse than random on average.

## Your own criterion, and the direction check

A user function receives the selected IDs and the data object and returns a number. Things
you want small (error, similarity) must be minimised; TSOpt probes the function and warns when
it looks like it is optimised in the wrong direction.

```{r custom-criterion}
similarity <- function(ids, data) mean(data$K[ids, ids])
warn <- tryCatch({ tso_design(tp, n = 30, criterion = similarity, iterations = 50, verbose = FALSE); "" },
                 warning = function(w) conditionMessage(w))
cat(substr(warn, 1, 120), "...\n")
chk("maximising a similarity triggers the wrong-direction warning", grepl("minimi", warn))
p_div <- tso_design(tp, n = 30, criterion = similarity, direction = "minimize", iterations = 100,
                    verbose = FALSE)
set.seed(1); rand_sim <- mean(replicate(20, similarity(sample(ids, 30), tp)))
c(minimised = similarity(p_div$selected$id, tp), random_sets = rand_sim)
chk("minimised similarity is below that of random sets", similarity(p_div$selected$id, tp) < rand_sim)
```

\newpage

# Every search

```{r s7, include = FALSE}
section <- "7 Searches"
```

The same criterion (CDmean at one heritability) optimised by each search. Greedy adds the best
line one at a time; exchange then swaps; ROTS (tabu) keeps moving past local optima; the genetic
algorithm with simulated annealing works for any criterion; restarts run the local searches from
several perturbed starts.

```{r searches}
run_search <- function(s, restarts = 0) {
  p <- timed(tso_design(tp, n = n_train, target = targets, h2 = params$h2, search = s,
                        restarts = restarts, iterations = 200, n_random = 20, verbose = FALSE))
  data.frame(search = paste0(s, if (restarts) paste0(" + ", restarts, " restarts") else ""),
             criterion_value = p$value$criterion_value %||% NA,
             exact_CDmean = tso_score(tp, train = p$selected$id, target = targets, h2 = params$h2,
                                      criteria = "cdmean_exact")[["cdmean_exact"]],
             seconds = round(attr(p, "seconds"), 2))
}
`%||%` <- function(a, b) if (is.null(a)) b else a
sr <- rbind(run_search("greedy"), run_search("greedy+exchange"), run_search("tabu"),
            run_search("tabu", restarts = 3), run_search("ga"))
sr[, 2:3] <- round(sr[, 2:3], 5)
knitr::kable(sr, row.names = FALSE)
cv <- sr$exact_CDmean
chk("exchange is at least as good as greedy", cv[2] >= cv[1] - 1e-6)
chk("tabu is at least as good as exchange", cv[3] >= cv[2] - 1e-6)
chk("restarts never make tabu worse", cv[4] >= cv[3] - 1e-6)
```

The search trace of the tabu search: the criterion after every accepted move. Greedy climbs
first, then exchange and tabu moves refine the set, sometimes accepting a worse move to escape a
local optimum:

```{r search-trace, fig.height = 3.4}
p_tabu <- tso_design(tp, n = n_train, target = targets, h2 = params$h2, search = "tabu", verbose = FALSE)
tso_plot(p_tabu, "search")
```

What to look at: the harder searches raise the criterion a little. In the benchmark this extra
criterion did **not** give better predictions, which is why the default is the plain greedy
search on robust CDmean.

\newpage

# Building training sets: the core options

```{r s8, include = FALSE}
section <- "8 Core designs"
```

## The default

Robust CDmean (averaged over heritability 0.2, 0.5 and 0.8), exact fixed intercept, greedy
search, automatic stratification when the panel has discrete groups.

```{r default}
plan <- tso_design(tp, n = n_train)
plan
chk("default returns n_train lines", nrow(plan$selected) == n_train)
chk("default beats every random set", plan$value$percentile >= 0.95,
    paste0("better than ", round(100 * plan$value$percentile), "% of random sets"))
chk("robust design reports CD at each heritability", length(plan$value$cd_by_h2) == 3)
```

**Reading the printout, line by line.**

* *Criterion*: what was optimised. "exact (fixed intercept)" means the trait mean is treated as
  an unknown fixed effect, as it is in the GBLUP you will fit; "averaged over h2 = 0.2, 0.5, 0.8"
  is the robust default.
* *Expected*: the CDmean (expected reliability of the target contrasts) of the chosen set,
  against the mean and spread of random sets of the same size, and the share of random sets it
  beats.
* *Verdict*: GREEN when the expected accuracy gain over random is 5% or more and the targets are
  covered, AMBER for 2 to 5%, RED below 2% (random sampling is then fine).
* *Structure*: whether the panel has discrete groups and how the lines were allocated to them.
* *Coverage*: the share of targets with a close relative (relationship of 0.20 or more) in the
  training set, and how many targets have no relative at all.
* *Divide*: candidates whose predictions will be too unreliable to select on.
* *Alleles*: the share of the targets' alleles and 5-marker haplotypes that also occur in the
  training set.

The selected lines, their expected reliability, and every unit's role:

```{r default-tables}
knitr::kable(utils::head(plan$selected[, intersect(c("order", "id", "group", "cost", "reliability"),
                                                    names(plan$selected))], 12), digits = 3)
table(plan$units$role)
```

```{r default-plots, fig.height = 3.4}
tso_plot(plan, "baseline")
tso_plot(plan, "pca")
```

## Targeted

```{r targeted}
plan_t <- tso_design(tp, n = n_train, target = targets, verbose = FALSE)
plan_t
chk("targets are never in the training set", !any(plan_t$selected$id %in% targets))
```

A fair comparison: the untargeted design above may phenotype target lines themselves, which
flatters it on those targets. Restricted to the same candidates (targets excluded), the targeted
design must predict the targets at least as well, at the heritability the score uses:

```{r targeted-fair}
plan_u <- tso_design(tp, n = n_train, candidates = setdiff(ids, targets), h2 = params$h2, verbose = FALSE)
plan_t5 <- tso_design(tp, n = n_train, target = targets, h2 = params$h2, verbose = FALSE)
sc <- c(untargeted = tso_score(tp, train = plan_u$selected$id, target = targets, h2 = params$h2)[["cdmean"]],
        targeted = tso_score(tp, train = plan_t5$selected$id, target = targets, h2 = params$h2)[["cdmean"]])
round(sc, 4)
chk("targeted design predicts the targets better than an untargeted one from the same candidates",
    sc[["targeted"]] >= sc[["untargeted"]] - 1e-6)
```

## Heritability: known value or your own prior grid

```{r h2}
p_h <- tso_design(tp, n = n_train, target = targets, h2 = params$h2, verbose = FALSE)
p_grid <- tso_design(tp, n = n_train, target = targets, h2 = c(0.1, 0.3, 0.5), verbose = FALSE)
round(p_grid$value$cd_by_h2, 4)
chk("a single h2 is used as given", isTRUE(all.equal(p_h$spec$h2, params$h2)))
```

## A budget with unit costs

```{r budget}
if (!is.null(costs)) {
  bud <- 2 * n_train * stats::median(tp$units$cost)
  p_bud <- tso_design(tp, budget = bud, target = targets, verbose = FALSE)
  c(budget = bud, spent = p_bud$spent, lines = nrow(p_bud$selected))
  chk("budget is respected", p_bud$spent <= bud + 1e-8)
} else cat("No cost file given: budget designs skipped.\n")
```

## Field reality: checks, lost seed and last season's data

```{r field-reality}
chk_lines <- ids[1:3]; lost <- ids[4:10]; last_year <- ids[11:40]
p_real <- tso_design(tp, n = n_train, must_include = chk_lines, exclude = lost, history = last_year,
                     history_weight = 0.8, verbose = FALSE)
head(p_real$selected[, c("order", "id", "forced")], 5)
chk("checks are always included", all(chk_lines %in% p_real$selected$id))
chk("excluded lines are never selected", !any(lost %in% p_real$selected$id))
chk("history lines are not selected again", !any(last_year %in% p_real$selected$id))
chk("history lines are never evaluation targets", !any(p_real$index$history %in% p_real$index$eval))
```

## Quotas, caps and replicates

```{r quotas}
if (!is.null(groups)) {
  grp_lev <- names(table(tp$units$group))
  q <- setNames(rep(floor(n_train / length(grp_lev)), length(grp_lev)), grp_lev)
  p_q <- tso_design(tp, n = sum(q), group_quota = q, verbose = FALSE)
  print(table(p_q$selected$group))
  chk("group quotas are met exactly", all(table(p_q$selected$group)[grp_lev] == q))
}
p_fam <- tso_design(tp_out, n = n_train, max_per_family = 3, verbose = FALSE)
chk("no family exceeds the cap of 3", max(table(p_fam$selected$family)) <= 3)
p_rep <- tso_design(tp, n = n_train, reps = 2, h2 = 0.2, stratify = "none", verbose = FALSE)
chk("replicated design fills all plots", nrow(p_rep$selected) == n_train)
chk("no line gets more than 2 plots", max(table(p_rep$selected$id)) <= 2)
```

## Fixed effects: trials, blocks, covariates

Fixed effects are handled exactly. Every level of a factor fixed effect is opened with two
units, because the first unit at a level only estimates that level's mean.

```{r fixed}
tp_tr <- tp; tp_tr$units$trial <- rep(c("A", "B", "C"), length.out = N)
p_fx <- tso_design(tp_tr, n = n_train, fixed = "trial", verbose = FALSE)
p_fx$spec$fixed
table(p_fx$selected$trial)
chk("every trial gets at least 2 lines", all(table(factor(p_fx$selected$trial, levels = c("A", "B", "C"))) >= 2))
cov <- matrix(rnorm(N), ncol = 1, dimnames = list(ids, "moisture"))
p_cov <- tso_design(tp, n = n_train, fixed = cov, verbose = FALSE)
chk("a covariate matrix is accepted as a fixed effect", identical(p_cov$spec$fixed_columns, "moisture"))
```

## Genetic model: total or breeding value

With a non-additive model, a design can predict the total genetic value (the line itself) or the
breeding value (the additive part, for choosing parents). Both are exact:

```{r value-bv}
tp_aa <- tso_data(g, kernels = c("A", "AA"), qc = FALSE, verbose = FALSE)
p_tv <- tso_design(tp_aa, n = n_train, target = targets, verbose = FALSE)
p_bv <- tso_design(tp_aa, n = n_train, target = targets, value = "breeding", verbose = FALSE)
p_bv
chk("the robust design averages over 3 share x 3 heritability scenarios", length(p_bv$value$cd_by_scenario) == 9)
Sb <- match(p_bv$selected$id, tp_aa$units$id); Tb <- setdiff(match(targets, tp_aa$units$id), Sb)
Cb <- diag(length(Tb)) - 1 / length(Tb); Kb <- tp_aa$K; Ka <- tp_aa$K_add
Vb <- Kb[Sb, Sb] + diag(1, length(Sb))
cd_manual <- mean(diag(Cb %*% Ka[Tb, Sb] %*% solve(Vb, Ka[Sb, Tb]) %*% Cb) / diag(Cb %*% Ka[Tb, Tb] %*% Cb))
chk("breeding-value CDmean matches the direct formula",
    abs(tso_score(tp_aa, p_bv$selected$id, target = targets, value = "breeding")[["cdmean"]] - cd_manual) < 1e-8)
chk("the plan reports that it predicts breeding values", identical(p_bv$spec$value, "breeding"))
cat("lines in common between the total-value and breeding-value plans:",
    length(intersect(p_tv$selected$id, p_bv$selected$id)), "of", n_train, "\n")
```

\newpage

# Training sets for every crop and breeding system

```{r s9, include = FALSE}
section <- "9 Breeding systems"
```

## Wheat, rice, barley: inbred lines, several goals

```{r goals}
goals <- c("accuracy", "gain", "information", "reference_panel", "correlation")
gres <- do.call(rbind, lapply(goals, function(gl) {
  p <- tso_design(tp, n = n_train, target = targets, goal = gl, iterations = 100, n_random = 20,
                  verbose = FALSE)
  data.frame(goal = gl, criterion = p$spec$criterion, lines = nrow(p$selected),
             exact_CDmean = round(tso_score(tp, train = p$selected$id, target = targets,
                                            criteria = "cdmean_exact")[["cdmean_exact"]], 4))
}))
knitr::kable(gres, row.names = FALSE)
chk("every goal returns a full training set", all(gres$lines == n_train))
```

## Maize, sorghum: hybrids

```{r hybrids}
p_hyb <- tso_design(tp_hyb, n = 3 * n_pool, max_per_parent = 6, verbose = FALSE)
p_hyb
par_use <- table(c(p_hyb$selected$parent1, p_hyb$selected$parent2))
chk("no parent is used in more than 6 training hybrids", max(par_use) <= 6)
chk("only crossable parents appear in training",
    all(p_hyb$selected$parent1 %in% fem[1:(n_pool - 5)] & p_hyb$selected$parent2 %in% mal[1:(n_pool - 5)]))
p_h0 <- tso_design(tp_hyb, n = 3 * n_pool, predict = "H0", verbose = FALSE)
chk("predict = H0 targets hybrids of two untested parents", grepl("H0", p_h0$spec$target))
```

```{r hybrid-plot, fig.height = 5}
tso_plot(p_hyb, "hybrid_grid")
```

## Potato (tetraploid) and cassava (diploid): clones

```{r clones}
p_pot <- tso_design(tp_potato, n = 40, verbose = FALSE)
p_cas <- tso_design(tp_clone, n = n_train, verbose = FALSE)
c(potato = p_pot$value$cdmean, cassava = p_cas$value$cdmean)
p_pot
chk("tetraploid clonal design runs", nrow(p_pot$selected) == 40)
```

```{r clones-plot, fig.height = 3.6}
tso_plot(p_pot, "pca")
```

```{r clones-checks}
chk("diploid clonal design runs", nrow(p_cas$selected) == n_train)
```

## Ryegrass, alfalfa: outcrossing families (CDpop)

```{r outcross-design}
p_pop <- tso_design(tp_out, n = n_train, goal = "families", verbose = FALSE)
chk("family-mean design (CDpop) runs", p_pop$spec$criterion == "cdpop")
length(unique(p_pop$selected$family))
```

## Genebank core collection

```{r genebank-design}
p_core <- tso_design(tp_bank, n = n_train, goal = "reference_panel", verbose = FALSE)
p_cov2 <- tso_design(tp_bank, n = n_train, criterion = "coverage", verbose = FALSE)
c(reference_panel = p_core$coverage$frac_020, coverage = p_cov2$coverage$frac_020)
chk("coverage design reaches most lines with a relative >= 0.20", p_cov2$coverage$frac_020 > 0.5)
```

## Rice, multi-location: sparse multi-environment testing

Every line is tested in one site only (`max_per_line = 1`), the noisy Site2 gets fewer plots,
and every site keeps at least 20.

```{r met-design}
p_met <- tso_design(tp_met, n = 3 * n_train, target = targets, max_per_line = 1, min_per_env = 20,
                    verbose = FALSE)
p_met
chk("every line is tested in one site only", all(table(p_met$selected$line_id) == 1))
chk("every site gets at least 20 lines", all(p_met$envs$allocation >= 20))
chk("target lines are never tested", !any(p_met$selected$line_id %in% targets))
p_met_def <- tso_design(tp_met, n = 3 * n_train, target = targets, verbose = FALSE)
chk("without constraints every site is still opened", all(p_met_def$envs$allocation >= 2))
```

```{r met-plot, fig.height = 4.5}
tso_plot(p_met, "met_allocation")
```

## Wheat quality: multi-trait index (which trait on which line)

```{r multitrait}
p_mt <- tso_design(tp_mt, budget = 40 * n_train, index = c(yield = 1, protein = 0.4, ndvi = 0),
                   verbose = FALSE)
p_mt$traits
chk("multi-trait budget is respected", p_mt$spent <= 40 * n_train + 1e-8)
```

## Very large panels

```{r large}
big <- tso_simulate_panel(n = 3000, markers = 1500, seed = 5)
p_big <- timed(tso_design_large(big, n = 200, rank = 200, verbose = FALSE))
p_big
chk("large-panel engine returns 200 lines", nrow(p_big$selected) == 200)
small <- tso_simulate_panel(n = 150, markers = 300, seed = 9)
lr <- tso_design_large(small, n = 30, h2 = 0.5, rank = 300, verbose = FALSE)
dn <- tso_design(small, n = 30, h2 = 0.5, search = "greedy", stratify = "none", verbose = FALSE)
chk("with exact features the large engine picks the same lines as the dense one",
    setequal(lr$selected$id, dn$selected$id))
```

\newpage

# Major genes and haplotypes

```{r s10, include = FALSE}
section <- "10 Major genes"
```

## Scanning your phenotypes

```{r pheno}
ph <- utils::read.csv(pheno_file)
y <- stats::setNames(ph$y, as.character(ph$id))
chk("phenotype IDs match the genotypes", mean(names(y) %in% ids) > 0.9,
    paste0(round(100 * mean(names(y) %in% ids)), "% matched"))
scan <- tso_major_scan(g, y)
```

```{r scan-plot, fig.height = 3.6}
tso_plot(scan)
```

## A trait with a planted major gene

To check that the machinery works, a trait with one major QTL (40% of the genetic variance)
is simulated on your genotypes:

```{r scan-planted}
tr_m <- tso_simulate_trait(g, h2 = 0.6, n_qtl = 80, major = 1, major_share = 0.4, seed = 11)
scan_m <- tso_major_scan(g, stats::setNames(tr_m$y, tr_m$id), verbose = FALSE)
scan_m
chk("the scan finds the planted major QTL", attr(tr_m, "major") %in% scan_m$leads$marker)
hp <- tso_haplotypes(g, around = attr(tr_m, "major"), window = 2)
hp
chk("haplotype block built around the major QTL", length(hp$blocks) == 1)
```

## Marker effects you already have

No phenotypes are needed when marker effects are known: from a GWAS, a QTL study, a BayesB
model or cloned genes. `tso_major_effects()` matches them to your markers and alleles and
computes each locus's share of genetic variance in your panel.

```{r effects-file}
if (nzchar(eff_file)) {
  me <- tso_major_effects(g, eff_file)
  chk("marker effects matched to the genotypes", me$n_markers > 0,
      paste0(me$n_markers, " of ", me$n_given, " markers"))
  chk("each lead's share is a proportion", all(is.na(me$leads$share_region)) ||
        all(me$leads$share_region >= 0 & me$leads$share_region <= 1))
  if (nrow(me$leads)) {
    tp_me <- tso_data(g, major = me, qc = FALSE, verbose = FALSE)
    chk("tso_data() protects the loci from the effects", setequal(tp_me$major$loci, me$leads$marker))
  }
} else cat("No marker-effects file given (params$effects).\n")
```

```{r effects-plot, fig.height = 3.2, eval = nzchar(eff_file)}
tso_plot(me)
```

The same function reads GWAS summary statistics. The scan of the planted trait stands in for a
published GWAS; the planted locus (40% of the genetic variance) must be found and recognised as
major. Its share will be too high: a GWAS effect estimated on the same few hundred lines is
inflated by the winner's curse (the Beavis effect), so the share from a small study is an upper
bound. Effects from a large published study, or the scan's joint estimate (`share_joint`), are
better guides to the size.

```{r effects-gwas}
gw <- data.frame(snp = scan_m$table$marker, beta = scan_m$table$effect, pval = 10^-scan_m$table$log10p)
me_gw <- tso_major_effects(g, gw, var_g = stats::var(tr_m$g))
chk("GWAS effects find the planted major QTL", attr(tr_m, "major") %in% me_gw$leads$marker)
sh_gw <- me_gw$leads$share_genetic[me_gw$leads$marker == attr(tr_m, "major")][1]
chk("the planted QTL is recognised as a major locus (share above 20%)", isTRUE(sh_gw > 0.2),
    paste0(round(100 * sh_gw), "% from the GWAS effect; scan's joint estimate ",
           round(100 * scan_m$leads$share_joint[scan_m$leads$marker == attr(tr_m, "major")][1]), "%"))
```

## Protecting the alleles in the training set

```{r protect}
tp_major <- tso_data(g, major = attr(tr_m, "major"), min_carriers = 3, qc = FALSE, verbose = FALSE)
p_major <- tso_design(tp_major, n = n_train, target = targets, verbose = FALSE)
p_major$major
chk("every allele class carried by targets has the required carriers",
    all(p_major$major$carriers_training >= p_major$major$required))
tp_fixed <- tso_data(g, major = list(markers = attr(tr_m, "major"), haplotypes = hp),
                     major_model = "fixed", qc = FALSE, verbose = FALSE)
p_fixed <- tso_design(tp_fixed, n = n_train, target = targets, verbose = FALSE)
chk("the fixed model uses the major loci as fixed effects", !is.null(p_fixed$spec$fixed_columns))
```

```{r protect-plot, fig.height = 3}
tso_plot(p_major, "major")
```

## Genome-wide allele and haplotype coverage

```{r allele-coverage}
unlist(plan_t$alleles[c("frac", "n_missing_markers")])
plan_t$alleles$haplotypes$frac
chk("allele coverage is a proportion", plan_t$alleles$frac >= 0 && plan_t$alleles$frac <= 1)
```

## Haplotype blocks, the haplotype kernel and haplotype coverage

Blocks of adjacent markers in LD (or fixed tiles, or windows in kb) define haplotypes. Your
material is `r if (all(g$dosage %in% c(0, g$ploidy))) "homozygous, so the marker strings are haplotypes" else if (isTRUE(g$phased)) "phased, so each chromosome copy gives a haplotype" else "heterozygous and unphased, so blocks hold local multi-locus genotypes (phase the file for true haplotypes)"`.

```{r haplotypes}
hp_ld <- tso_haplotypes(g, method = "ld")
hp_fx <- tso_haplotypes(g, method = "fixed", window = 2)
hp_ld
chk("LD blocks use every marker exactly once", setequal(unlist(hp_ld$blocks), colnames(g$dosage)) &&
      !anyDuplicated(unlist(hp_ld$blocks)))
chk("haplotype frequencies sum to 1 in every block", all(abs(vapply(hp_fx$freq, sum, 1) - 1) < 1e-10))
hp1 <- tso_haplotypes(g, method = "fixed", window = 0, min_freq = 0)
H1 <- TSOpt:::.kernel_hap(hp1)
g1 <- g                                            # haplotypes use rounded dosages (homozygous lines: 0 or ploidy)
g1$dosage <- if (hp1$kind == "homozygous") g$ploidy * round(round(g$dosage) / g$ploidy) else round(g$dosage)
A1 <- tso_kinship(g1, base_freq = colMeans(g1$dosage) / g$ploidy, ridge = 0)
chk("single-marker blocks reproduce the VanRaden matrix exactly", max(abs(H1 - A1 / mean(diag(A1)))) < 1e-8)
tp_h <- tso_data(g, haplotypes = hp_fx, qc = FALSE, verbose = FALSE)
p_def <- tso_design(tp_h, n = n_train, target = targets, verbose = FALSE)
p_hap <- tso_design(tp_h, n = n_train, target = targets, criterion = "haplotype", verbose = FALSE)
cov_tab <- rbind(default = unlist(p_def$alleles$haplotypes[c("frac", "frac_copies", "frac2")]),
                 haplotype = unlist(p_hap$alleles$haplotypes[c("frac", "frac_copies", "frac2")]))
round(cov_tab, 3)
chk("the haplotype criterion covers at least as many target haplotype copies",
    cov_tab["haplotype", "frac_copies"] >= cov_tab["default", "frac_copies"] - 1e-12)
chk("the per-block report has one row per block", nrow(p_hap$alleles$haplotypes$per_block) == length(hp_fx$blocks))
```

```{r haplotypes-plot, fig.height = 3.4}
tso_plot(p_def, "haplotypes")
```

\newpage

# Checking a plan

```{r s11, include = FALSE}
section <- "11 Check"
```

## Why this set: tso_explain()

```{r explain}
tso_explain(plan_t, steps = FALSE)
```

## Any set, any criterion: tso_score()

```{r score}
set.seed(2); colleague <- sample(setdiff(ids, targets), n_train)
rbind(colleague = tso_score(tp, train = colleague, target = targets),
      optimised = tso_score(tp, train = plan_t$selected$id, target = targets))
chk("the optimised set scores higher than a random colleague's set",
    tso_score(tp, train = plan_t$selected$id, target = targets)[["cdmean"]] >
      tso_score(tp, train = colleague, target = targets)[["cdmean"]])
```

## Against the alternatives: tso_compare()

```{r compare}
cmp <- tso_compare(plan_t, others = list(colleague = colleague), n_random = 50)
cmp
chk("the plan is at least as good as random sampling",
    cmp$table$cdmean[cmp$table$method == "optimised"] >= cmp$table$cdmean[cmp$table$method == "random"])
```

```{r compare-plot, fig.height = 3.2}
tso_plot(cmp)
```

## Top-k identification: tso_topk()

```{r topk}
tk <- tso_topk(plan_t, top = 0.1, n_sim = 300)
tk
chk("top-k analysis reports plan and random-set metrics", is.data.frame(tk$summary) && nrow(tk$summary) > 0)
```

```{r topk-plot, fig.height = 3.6}
tso_plot(tk)
```

## Accuracy against diversity: tso_pareto()

```{r pareto}
fr <- tso_pareto(tp, n = n_train, objectives = c("cdmean", "coancestry"), target = targets,
                 pop = 50, generations = 50, verbose = FALSE)
fr
knee <- tso_pareto_pick(fr, "knee")
chk("the knee set has n_train lines", length(knee) == n_train)
chk("the front has more than one set", nrow(fr$front) > 1)
```

```{r pareto-plot, fig.height = 4}
tso_plot(fr)
```

\newpage

# After the season: evaluate the plan

```{r s12, include = FALSE}
section <- "12 Evaluate"
```

Once the training set is phenotyped, `tso_evaluate()` fits GBLUP on it and reports how well it
predicted the targets: predictive ability, rank correlation, NDCG, realised gain from selecting
the top 10% and the bias slope. With true genetic values (simulations) the metrics are computed
against the truth.

```{r evaluate}
tv <- if (nzchar(truth_file) && file.exists(truth_file)) {
  d <- utils::read.csv(truth_file); stats::setNames(d$g, as.character(d$id)) } else NULL
ev <- tso_evaluate(plan_t, y, truth = tv, h2 = params$h2)
ev
chk("predictive ability is a correlation", abs(ev$predictive_ability) <= 1)
chk("evaluation never scores the training lines", !any(ev$predictions$id %in% plan_t$selected$id))
set.seed(3)
rnd <- sample(setdiff(ids, targets), n_train)
ev_r <- tso_evaluate(tp, y, train = rnd, target = targets, truth = tv, h2 = params$h2)
c(optimised = ev$predictive_ability, random = ev_r$predictive_ability)
```

```{r evaluate-plot, fig.height = 3.8}
tso_plot(ev)
```

## Real accuracy of every criterion

Section 6 compared the criteria on an expected quantity. Here each criterion's training set is
phenotyped (the phenotypes are in `y`) and GBLUP predicts the targets. Against the true genetic
values when they are known (simulated data), otherwise against the targets' own phenotypes.
This is the measure that matters to a breeder.

```{r criteria-realised}
real <- do.call(rbind, lapply(names(crit_sets), function(cr) {
  e <- tso_evaluate(tp, y, train = crit_sets[[cr]], target = targets, truth = tv, h2 = params$h2)
  data.frame(criterion = cr, predictive_ability = e$predictive_ability, rank_corr = e$spearman,
             ndcg = e$ndcg, realised_gain_sd = e$realised_gain)
}))
real <- real[order(-real$predictive_ability), ]
knitr::kable(real, digits = 3, row.names = FALSE)
chk("every criterion's set can be evaluated", nrow(real) == length(crit_sets))
set.seed(5)
rand_pa <- mean(replicate(10, tso_evaluate(tp, y, train = sample(setdiff(ids, targets), n_train),
                                           target = targets, truth = tv, h2 = params$h2)$predictive_ability))
cat("mean predictive ability of 10 random training sets:", round(rand_pa, 3), "\n")
chk("the default cdmean predicts at least as well as the average random set",
    real$predictive_ability[real$criterion == "cdmean"] >= rand_pa - 0.02)
```

What to look at: one panel and one trait give a noisy ranking (a single random set can be
lucky), which is why the package default rests on a benchmark over six panels and 33
scenarios. On your own data, a criterion near the bottom of this table and near the top of the
Section 6 table has been optimising the wrong quantity for this trait.

```{r criteria-realised-plot, fig.height = 4.2}
ggplot(real, aes(stats::reorder(criterion, predictive_ability), predictive_ability)) +
  geom_col(fill = tso_colors()[["training"]]) + coord_flip() + tso_theme() +
  labs(x = NULL, y = "predictive ability on the targets", title = "Realised accuracy of every criterion")
```

## Unreliable heritability estimates

Without a heritability, REML estimates it from the training set. With few lines the estimate
can hit its boundary, and TSOpt then warns that the bias slope and realised gain are unreliable:

```{r evaluate-reml}
ev_reml <- withCallingHandlers(tso_evaluate(plan_t, y, truth = tv), warning = function(w) {
  cat("TSOpt warns:", substr(conditionMessage(w), 1, 100), "...\n"); invokeRestart("muffleWarning") })
ev_reml$h2_used
```

\newpage

# Across stages, seasons and generations

```{r s13, include = FALSE}
section <- "13 Seasons"
```

## Second stage: tso_next_batch()

```{r next-batch}
y1 <- y[intersect(plan$selected$id, names(y))]
nb <- tso_next_batch(plan, y1, n = 25, select_top = 0.1, h2 = params$h2)
nb
chk("next batch has 25 plots", nrow(nb$selected) == 25)
chk("posterior probabilities of selection lie in [0, 1]",
    all(nb$posterior$p_top >= 0 & nb$posterior$p_top <= 1, na.rm = TRUE))
```

```{r next-batch-plot, fig.height = 3.8}
tso_plot(nb)
```

## Next season: tso_update()

```{r update}
p_next <- tso_update(plan, data = tp, n = 40, history_weight = 0.8, verbose = FALSE)
p_next$spec$history
chk("last season's lines become history, not new plots", !any(plan$selected$id %in% p_next$selected$id))
```

## Choosing parents: tso_contribution()

```{r ocs}
ebv <- stats::setNames(ev$predictions$predicted, ev$predictions$id)
ocs <- tso_contribution(ebv, tp$K_lines[names(ebv), names(ebv)], delta_F = 0.01)
ocs
chk("contributions sum to 1", abs(sum(ocs$contributions$contribution) - 1) < 1e-8)
chk("contributions are non-negative", all(ocs$contributions$contribution >= -1e-12))
chk("the coancestry limit is respected", ocs$coancestry <= ocs$limit + 1e-8)
```

```{r ocs-plot, fig.height = 3.6}
tso_plot(ocs)
```

## The long term: tso_simulate()

```{r simulate}
sim <- tso_simulate(strategies = list(cdmean = "cdmean", random = "random"), cycles = 3, reps = 1,
                    n_candidates = 120, n_train = 40, n_parents = 12, markers = 500, verbose = FALSE)
sim
chk("the simulation reports every cycle and strategy", nrow(sim$results) == 3 * 2)
```

```{r simulate-plot, fig.height = 4.6}
tso_plot(sim)
```

\newpage

# Decisions around the design

```{r s135, include = FALSE}
section <- "13b Decisions around the design"
```

## Traits and environments together, hedged over uncertain correlations

A unit is now one trait on one line in one environment. The design hedges the index over
uncertain heritabilities and genetic correlations, and prints the CDmean under every scenario.

```{r mt-env}
g150 <- g; g150$dosage <- g$dosage[seq_len(min(150, N)), ]
tp_mte <- tso_data(g150, traits = c("yield", "protein", "ndvi"), trait_h2 = c(0.3, 0.5, 0.6),
                   gcor = matrix(c(1, .5, .6, .5, 1, .3, .6, .3, 1), 3),
                   trait_cost = c(yield = 20, protein = 35, ndvi = 0.5), plot_cost = 10,
                   envs = c("Site1", "Site2"), env_cor = matrix(c(1, .6, .6, 1), 2),
                   env_noise = c(Site1 = 1, Site2 = 2), qc = FALSE, verbose = FALSE)
p_mte <- tso_design(tp_mte, budget = 2500, index = c(yield = 1, protein = 0.4), verbose = FALSE)
p_mte
table(p_mte$selected$trait, p_mte$selected$env)
chk("line x trait x environment units", nrow(tp_mte$units) == nrow(g150$dosage) * 6)
chk("the multi-trait budget across sites is respected", p_mte$spent <= 2500 + 1e-8)
chk("the robust multi-trait design reports every scenario", length(p_mte$value$cd_by_scenario) >= 4)
chk("every trait is measured at least twice (trait means are estimable)",
    all(table(factor(p_mte$selected$trait, levels = c("yield", "protein", "ndvi"))) >= 2))
```

## What is each trait worth?

```{r trait-value}
tp_tv <- tso_data(g, traits = c("yield", "protein", "ndvi"), trait_h2 = c(0.3, 0.5, 0.6),
                  gcor = matrix(c(1, .5, .6, .5, 1, .3, .6, .3, 1), 3),
                  trait_cost = c(yield = 20, protein = 35, ndvi = 0.5), plot_cost = 10, qc = FALSE,
                  verbose = FALSE)
p_tv <- tso_design(tp_tv, budget = 3000, index = c(yield = 1, protein = 0.4), verbose = FALSE)
tv <- tso_trait_value(p_tv)
tv
chk("budget shares sum to 1", abs(sum(tv$table$share_of_budget) - 1) < 1e-8)
chk("the reference trait has exchange rate 1", abs(tv$table$exchange_rate[tv$table$trait == "yield"] - 1) < 1e-8)
```

```{r trait-value-plot, fig.height = 3}
tso_plot(tv)
```

## The value of each plot

```{r plot-value}
knitr::kable(utils::head(plan_t$selected[order(-plan_t$selected$value_pct), c("id", "value_pct")], 5),
             digits = 2, row.names = FALSE)
K_t <- tp$K; Ke_t <- K_t + 1e3 * mean(diag(K_t)); tix <- match(targets, ids)
ctr_t <- TSOpt:::.contrasts(K_t, tix, "centered")
s2_t <- rep(max((1 - p_h$spec$h2) / p_h$spec$h2, 1e-6), N)
cdv <- function(S) { st <- TSOpt:::.engine_init(Ke_t, ctr_t); for (j in S) TSOpt:::.engine_add(st, j, s2_t[j])
  TSOpt:::.crit_value("cdmean", st) }
S_h <- p_h$index$selected
chk("plot value equals the brute-force loss of dropping the plot (first 5 plots)",
    max(abs(sapply(1:5, function(i) cdv(S_h) - cdv(S_h[-i])) - p_h$selected$value[1:5])) < 1e-8)
```

```{r plot-value-plot, fig.height = 4.5}
tso_plot(plan_t, "value")
```

## How sure is the advantage? tso_uncertainty()

```{r uncertainty}
unc <- tso_uncertainty(plan_t, B = 30)
unc
chk("the uncertainty interval contains the estimate",
    all(unc$intervals$lower <= unc$intervals$estimate & unc$intervals$estimate <= unc$intervals$upper))
```

```{r uncertainty-plot, fig.height = 3}
tso_plot(unc)
```

## How many plots pay off? tso_economics()

```{r economics}
eco <- tso_economics(tp, cost_per_plot = 40, value_per_sd = 5000, select_top = 0.1, target = targets)
eco
chk("the economic optimum maximises net value on the curve",
    abs(eco$optimum$opt$net - max(eco$curve$net_opt, na.rm = TRUE)) < 1e-6)
```

```{r economics-plot, fig.height = 3.4}
tso_plot(eco)
```

## Laying out the field: tso_layout()

```{r layout}
lay <- tso_layout(plan_t, rows = 10, cols = ceiling((n_train + 6) / 10) + 1, blocks = 2,
                  checks = ids[1:2], check_reps = 1, prep = 0.1)
lay
fld <- as.data.frame(lay)
chk("every training line is planted", all(plan_t$selected$id %in% fld$id))
chk("blocking never raises the CDmean", lay$metrics$cd_blocks <= lay$metrics$cd_no_blocks + 1e-8)
```

```{r layout-plot, fig.height = 4}
tso_plot(lay)
```

## Whom to genotype, and hedging over the model of gene action

```{r genotype-kernels}
gs <- tso_genotype_set(tp, n = 20)
gs
chk("the genotyped set beats random sets on coverage",
    gs$comparison$mean_best[1] >= gs$comparison$mean_best[nrow(gs$comparison)] - 1e-8 ||
      gs$percentile >= 0.5)
p_kern <- tso_design(tp, n = n_train, target = targets, kernels = c("additive", "gaussian"), h2 = params$h2,
                     verbose = FALSE)
round(p_kern$value$cd_by_scenario, 4)
chk("the kernel-robust design reports every kernel", length(p_kern$value$cd_by_scenario) == 2)
```

## Training in one programme, predicting another: tso_transfer()

```{r transfer}
if (!is.null(groups) && length(unique(tp$units$group)) >= 2) {
  grp_names <- sort(unique(tp$units$group))
  tr_out <- tso_transfer(tp, from = grp_names[-length(grp_names)], to = grp_names[length(grp_names)],
                         n = n_train, n_local = 10)
  tr_out
  chk("transferability is a ratio between 0 and 1", tr_out$transferability > 0 && tr_out$transferability <= 1.05)
} else cat("No groups: transfer skipped.\n")
```

## Straight from a breeding database (BrAPI)

With a real server you connect and read genotypes and phenotypes directly:

```{r brapi-live, eval = FALSE}
con <- tso_brapi_connect("https://my-breedbase.org", token = Sys.getenv("BRAPI_TOKEN"))
tso_brapi_studies(con)
g_db <- tso_brapi_geno(con, variant_set = "vs1")
y_db <- tso_brapi_trait(con, study = "trial2025", trait = "grain yield")
tp_db <- tso_data(g_db)
```

Here the connector is checked offline against the recorded BrAPI responses that ship with the
package's tests, when the repository is available:

```{r brapi-offline}
fx <- normalizePath(file.path("..", "..", "tests", "testthat", "fixtures", "brapi"), mustWork = FALSE)
if (dir.exists(fx)) {
  con <- tso_brapi_connect(paste0("file://", fx), verbose = FALSE)
  g_db <- tso_brapi_geno(con, verbose = FALSE)
  g_db
  chk("BrAPI genotypes are read into a tso_geno", inherits(g_db, "tso_geno") && nrow(g_db$dosage) > 0)
} else cat("BrAPI fixtures not found (run from the repository): offline check skipped.\n")
```

\newpage

# Programme planning and hand-over

```{r s-prog, include = FALSE}
section <- "Programme"
```

## Variance components from your phenotypes: tso_variance()

```{r prog-variance}
v_y <- tso_variance(tp, data.frame(id = names(y), y = y))
v_y
chk("REML converged", isTRUE(v_y$converged))
chk("the heritability is a proportion", all(v_y$h2 > 0 & v_y$h2 < 1))
chk("the same estimate as the GBLUP fit", abs(v_y$h2 - TSOpt:::.gblup(tp$K_lines, match(names(y)[names(y) %in% rownames(tp$K_lines)],
    rownames(tp$K_lines)), y[names(y) %in% rownames(tp$K_lines)])$h2) < 0.01)
tp_v <- tso_data(g, variance = v_y, qc = FALSE, verbose = FALSE)
chk("robust designs use the estimated heritability grid", length(tp_v$h2_grid) >= 1 &&
      all(tp_v$h2_grid > 0 & tp_v$h2_grid < 1))
```

## Forward validation: tso_backtest()

```{r prog-backtest}
bt_hist <- if (nzchar(params$history)) utils::read.csv(params$history, stringsAsFactors = FALSE) else NULL
if (!is.null(bt_hist)) {
  bt <- tso_backtest(tp, bt_hist, n = 0.2, traits = params$history_trait, methods = c("random", "default", "kinship"),
                     n_random = 3, verbose = FALSE)
} else {
  hs <- tso_simulate_cycles(cycles = 4, n_candidates = 100, n_crosses = 10, n_parents = 10, markers = 500, seed = 4)
  bt <- tso_backtest(hs$geno, hs$pheno, n = 0.25, methods = c("random", "default", "kinship"), n_random = 3,
                     verbose = FALSE)
}
bt
chk("every test cycle and method is scored", nrow(bt$results) == length(unique(bt$results$cycle)) * 3)
chk("predictive abilities are correlations", all(abs(bt$results$pa) <= 1))
```

```{r prog-backtest-plot, fig.height = 3.2}
tso_plot(bt)
```

## The season across stages: tso_season()

```{r prog-season}
sea <- tso_season(tp, budget = round(0.6 * N), n1 = unique(round(N * c(0.1, 0.2, 0.3))),
                  k1 = unique(pmax(5, round(N * c(0.03, 0.06)))), sites2 = 1:3, reps = 10, verbose = FALSE)
sea
chk("every allocation fits the budget", all(sea$table$cost <= round(0.6 * N)))
chk("the best allocation has the highest expected gain", sea$best$gain == max(sea$table$gain))
```

## Which crosses to make: tso_crosses()

```{r prog-crosses}
has_map <- !is.null(g$map$cM) || params$cm_per_mb > 0
if (has_map) {
  parents <- names(sort(y[names(y) %in% ids], decreasing = TRUE))[1:10]
  cr <- tso_crosses(g, parents = parents, pheno = y, cM_per_Mb = if (params$cm_per_mb > 0) params$cm_per_mb else NULL)
  cr
  chk("45 crosses among 10 parents", nrow(cr$table) == 45)
  i10 <- stats::dnorm(stats::qnorm(0.9)) / 0.1
  chk("usefulness = mean + i x SD", max(abs(cr$table$usefulness - cr$table$mean - i10 * cr$table$sd)) < 1e-10)
} else cat("No genetic map: give params$cm_per_mb (map density of your crop) to predict crosses.\n")
```

## Genotyping, imputation and plots: tso_imputation_plan()

```{r prog-imputation}
M <- ncol(g$dosage)
ipl <- tso_imputation_plan(tp, budget = 40 * N, cost_hd = 40, cost_ld = 12, plot_cost = 25,
                           n_ref = unique(round(N * c(0.1, 0.25))), ld_markers = unique(round(M * c(0.05, 0.2))))
ipl
chk("every plan fits the budget", all(ipl$table$genotyping + 25 * ipl$table$plots <= 40 * N))
chk("realised reliability equals the expected one when nothing is imputed",
    with(ipl$table[ipl$table$plan == "all high density", , drop = FALSE],
         !length(reliability) || abs(reliability - cdmean_expected) < 1e-10))
```

## Testcrosses and the factorised form on your panel

```{r prog-testcross}
tcd <- tso_data(g, system = "testcross", testers = ids[1:2], qc = FALSE, verbose = FALSE)
chk("testcross units are lines x testers", nrow(tcd$units) == (N - 2) * 2)
ptc <- tso_design(tcd, n = min(40, N - 2), max_per_line = 1, verbose = FALSE)
chk("the testcross plan predicts GCA", identical(ptc$spec$value, "gca"))
sub <- ids[seq_len(min(80, N))]; gs <- g; gs$dosage <- g$dosage[sub, ]
e0 <- tso_data(gs, envs = c("E1", "E2", "E3"), qc = FALSE, verbose = FALSE)
e1 <- tso_data(gs, envs = c("E1", "E2", "E3"), qc = FALSE, implicit = TRUE, verbose = FALSE)
pe0 <- tso_design(e0, n = 30, h2 = 0.5, verbose = FALSE); pe1 <- tso_design(e1, n = 30, h2 = 0.5, verbose = FALSE)
chk("the factorised form gives the same design as the dense one",
    identical(pe0$selected$id, pe1$selected$id) && abs(pe0$value$cdmean - pe1$value$cdmean) < 1e-10)
```

## Omics profiles: tso_omics_kernel() and tso_omics_plan()

```{r prog-omics}
if (nzchar(params$omics)) {
  ok_all <- tso_omics_kernel(params$omics, type = params$omics_type, geno = tp)
} else {
  om_sim <- tso_simulate_omics(g, features = 100, seed = 3)       # simulated profiles of your lines
  ok_all <- tso_omics_kernel(om_sim$profiles, geno = tp)
}
ok_all
chk("the omics kernel has mean diagonal 1", abs(mean(diag(ok_all$K)) - 1) < 1e-10)
chk("the omics kernel is positive semi-definite", min(eigen(ok_all$K, TRUE, TRUE)$values) > -1e-8)
prof <- intersect(ok_all$lines, ids)
half <- prof[seq_len(max(3, floor(length(prof) / 2)))]
ok_half <- ok_all; ok_half$K <- ok_all$K[half, half]; ok_half$lines <- half
ok_half$blend <- 0          # no genomic blend (default 5%), so the profiled block is the measured one up to scale
tp_om <- tso_data(g, kernels = ok_half, qc = FALSE, verbose = FALSE)
chk("lines without a profile are kept (predicted from genomics)", nrow(tp_om$K) == N)
off <- upper.tri(ok_half$K)
chk("the profiled block keeps the measured relationships",
    stats::cor(tp_om$kernels$K[[ok_half$name]][half, half][off], ok_half$K[off]) > 0.9999)
pl_om <- tso_design(tp_om, n = min(params$n_train, N - 10), verbose = FALSE)
chk("the omics-aware design reports genetic + omics values",
    any(grepl("genetic \\+ omics", utils::capture.output(print(pl_om)))))
if (length(prof) >= 40) {
  opl <- tso_omics_plan(tp, ok_all, budget = 30 * length(prof), profile_cost = 2, plot_cost = 40,
                        n_profile = unique(round(length(prof) * c(0.25, 0.5))), strategy = c("coverage", "targets"))
  opl
  chk("profiling everyone: the realised reliability is the expected one",
      with(opl$table[opl$table$strategy == "all", , drop = FALSE], !length(realised) || abs(realised - expected) < 1e-10))
} else cat("Fewer than 40 profiled lines: no profiling plan.\n")
```

## To the field and the database

```{r prog-export}
fb <- tso_export_field(plan_t, format = "fieldbook", trial = "WALK")
chk("the Field Book file has one row per plot", nrow(fb) == nrow(plan_t$selected))
chk("FielDHub entries are unique", !anyDuplicated(tso_export_field(plan_t, format = "fieldhub")$NAME))
chk("the plan's manifest matches its data", all(tso_verify(plan_t, tp)$match))
js <- tempfile(fileext = ".json"); tso_plan_json(plan_t, js)
chk("the plan file reads back", identical(jsonlite::fromJSON(js)$selected$id, plan_t$selected$id))
code <- tso_analysis_script(plan_t, engine = "sommer")
chk("the analysis script parses", !inherits(try(parse(text = code), silent = TRUE), "try-error"))
```

\newpage

# Figures, export and reports

```{r s14, include = FALSE}
section <- "14 Outputs"
```

## Every plan figure

```{r fig-types}
types <- c("summary", "pca", "divide", "coverage", "baseline", "search", "kinship")
made <- sapply(types, function(ty) inherits(tso_plot(plan_t, ty), c("ggplot", "patchwork")))
made
chk("every plan figure type is drawn", all(made))
```

```{r fig-summary, fig.height = 6.5}
tso_plot(plan_t)
```

```{r fig-divide-coverage, fig.height = 3.4}
tso_plot(plan_t, "divide")
tso_plot(plan_t, "coverage")
```

## Journal-ready files, the field list and the report

```{r outputs}
f <- tso_save_figure(tso_plot(plan_t, "pca"), file.path(out, "fig_pca"), journal = "g3",
                     width = "single", format = c("pdf", "png", "tiff", "svg"), dpi = 300)
basename(f)
chk("all four figure formats were written", all(file.exists(f)))
chk("unknown journal presets are refused",
    inherits(try(tso_save_figure(tso_plot(plan_t, "pca"), file.path(out, "x"), journal = "nope"),
                 silent = TRUE), "try-error"))
csv <- tso_export(plan_t, file.path(out, "plan.csv"))
chk("the field list has one row per training line", nrow(utils::read.csv(file.path(out, "plan.csv"))) == n_train)
rep_file <- tso_report(plan_t, file.path(out, "report.html"), comparison = cmp)
chk("the HTML report is written", file.exists(rep_file) && file.size(rep_file) > 10000)
```

The theme and colours in your own figures:

```{r own-figure, fig.height = 3}
ggplot(plan_t$selected, aes(reliability)) +
  geom_histogram(bins = 20, fill = tso_colors()[["training"]]) + tso_theme() +
  labs(title = "Expected reliability of the selected lines", x = "reliability", y = "lines")
```

\newpage

# Robustness: inputs that should fail clearly

```{r s15, include = FALSE}
section <- "15 Robustness"
```

A good tool fails loudly and clearly on impossible requests instead of returning something
misleading. Each line below must produce an error or a warning.

```{r robustness}
err <- function(expr) inherits(try(expr, silent = TRUE), "try-error")
wrn <- function(expr) { w <- FALSE; withCallingHandlers(try(expr, silent = TRUE),
  warning = function(x) { w <<- TRUE; invokeRestart("muffleWarning") }); w }
chk("n larger than the pool is refused", err(tso_design(tp, n = N + 1, verbose = FALSE)))
chk("neither n nor budget is refused", err(tso_design(tp, verbose = FALSE)))
chk("unknown target IDs are refused", err(tso_design(tp, n = 10, target = "no_such_line", verbose = FALSE)))
chk("must_include larger than n is refused", err(tso_design(tp, n = 3, must_include = ids[1:6], verbose = FALSE)))
chk("an unknown criterion is refused", err(tso_design(tp, n = 10, criterion = "magic", verbose = FALSE)))
chk("h2 outside (0, 1] is refused", err(tso_design(tp, n = 10, h2 = 1.5, verbose = FALSE)))
chk("a constant custom criterion is refused",
    err(tso_design(tp, n = 10, criterion = function(i, d) 1, iterations = 10, verbose = FALSE)))
one_group <- if (!is.null(groups)) ids[tp$units$group == tp$units$group[1]] else NULL
if (length(one_group) > 5)
  chk("CDpop with targets from one group gives a clear error",
      err(tso_design(tp, n = 10, target = one_group[1:5], goal = "families", verbose = FALSE)))
chk("a single target works (no crash)",
    !err(tso_design(tp, n = 10, target = ids[1], verbose = FALSE)))
if (!is.null(groups)) {
  lev <- names(table(tp$units$group))
  chk("an unmeetable quota warns instead of silently shrinking",
      wrn(tso_design(tp, n = 3 * length(lev) + 10, group_quota = setNames(rep(3, length(lev)), lev),
                     verbose = FALSE)))
}
chk("history and must_include overlap is resolved (no duplicates)",
    !anyDuplicated(tso_design(tp, n = 10, history = ids[1:5], must_include = ids[3:7],
                              verbose = FALSE)$index$selected))
```

\newpage

# Scorecard

```{r s16, include = FALSE}
section <- "16 Scorecard"
```

```{r scorecard}
summary_tab <- aggregate(result ~ section, checks, function(r) sprintf("%d / %d", sum(r == "PASS"), length(r)))
names(summary_tab)[2] <- "passed"
knitr::kable(summary_tab, row.names = FALSE)
cat("\nTotal:", sum(checks$result == "PASS"), "of", nrow(checks), "checks passed.\n")
fails <- checks[checks$result == "FAIL", ]
if (nrow(fails)) knitr::kable(fails, row.names = FALSE) else cat("No check failed: nothing looks odd.\n")
```

Every check, in order:

```{r scorecard-all}
knitr::kable(checks, row.names = FALSE)
```

```{r session}
sessionInfo()
```
